---
title: "Tutorial on Pivot Tables and other Excel tasks you can also do in R - Witch Trials"
author: "Kim Fitter"
output: 
  html_document: 
    toc: true
    toc_float: true
---

Which witches and other Excel tasks.....in R. This tutorial assuming you have already installed R and R studio and had a brief introduction to the R basics. Run each line of code yourself or Knit to HTML to walk through this tutorial.

At the end of the exercise we will have an interactive tabular view of the number of witches tried by country.

```{r setup, include=FALSE}
# This is a comment on the setup chunk, we are using echo=TRUE so that the code is included in the R Markdown document
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```

## Particularly useful packages for this exercise

```{r install.packages}
# Install the package the first time you use the package
# install.packages("tidyverse")
# install.packages("rpivotTable")
library(tidyverse)
library(rpivotTable)
```

## Import Witch Trial data

The dataset used in this tutorial is of 43,000 people tried for witchcraft across
21 European countries over a period of five-and-a-half centuries used for this [economic journal](https://www.researchgate.net/publication/315533575_Witch_Trials). 

First import the data using the read_csv function from the [readr](https://cran.r-project.org/web/packages/readr/index.html) R package which is loaded with the tidyverse, to import the raw data to a dataframe called **witchdat**.

This function will import to a data frame, which is a 2-dimensional object where contents can be different types, as default.

```{r download data}
# Download data from the author's GitHub url
url <- "https://raw.githubusercontent.com/JakeRuss/witch-trials/master/data/trials.csv"
# The read_csv function also produces default messages describing the parsing of columns which is converting the columns into different types.
witchdat <- read_csv(url)
```
  
Let's take a quick look at the **witchdat** data frame. 

In R there are some packages that are included by default; take a look at the packages available in the Global Environment drop down. One such package is called package is [base](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html) which includes dim function to retrieve the dimensions, the row and column numbers.

```{r dim}
# Retrieve the number of rows and columns
dim(witchdat)
```

This is a fairly large dataframe!

Another included package is [utils](https://cran.r-project.org/web/packages/R.utils/index.html) R package which includes the str and structure functions to view the basic structure. 

```{r structure}
# Use str function to 'Compactly Display the Structure of an Arbitrary R Object'. All R objects store additional attributes to store metadata about an object. In this str function we set the give.attr to FALSE to exclude these attributes.
str(witchdat, give.attr=FALSE)
# Try running the command without the parameter...
# str(witchdat)
# Next we will use structure function to view witchdat 
structure(witchdat)
```
It looks like there are character variables and integer variables.

##  Format "cells" in R

In Excel we can format "cells". Which functions can we use in R?

A tool for data frame data manipulation is the  [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) R package. R packages typically have vignettes which are tutorials and worked examples using the package. There is useful documentation in the [Introduction to dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html).

In order to access the variables we can use a **$**. Alternatively we can use another method that joins objects and functions using a pipe **%>%**. This can be useful if we would like to use multiple dplyr verbs or functions on an object.

```{r format}
# Use dplyr to rename a variable
witchdat <- witchdat %>% 
  rename(country=gadm.adm0)
# Perform data type conversion of the deaths variable - use dplyr mutate to change 1 variable
witchdat <- witchdat %>% 
  mutate(deaths=as.numeric(deaths))
# or we can use dplyr mutate_all to change types of all integer variables to numeric
witchdat <- witchdat %>% 
  mutate_if(is.integer,as.numeric)
# View the structure of witchdat again to see these formatting changes
str(witchdat, give.attr=FALSE)
```

Try typing 'as.' in the console to see what other options are available to change data types.

For other formatting we can also format the end products (such as plots) that we create, using the functions in those R packages.

##  "Sort" data in R

In Excel we can sort by columns. Which functions can we use in R?

```{r arrange}
# Sort by character using arrange from dplyr. We can also pipe other packages functions such as head from the utils package.
witchdat %>% 
  arrange(city) %>% 
  head()
# This sorting also works on numeric variables, sort by year using arrange from dplyr
witchdat %>% 
  arrange(year) %>% 
  head()
# Or we can sort in descending order
witchdat %>% 
  arrange(desc(year)) %>% 
  head()
```

##  Create a "formula" in R

In Excel we can create formulas in cells. Which functions can we use in R?

```{r formula}
# Create a new variable in our dataframe with mutate from dplyr. 
witchdat <- witchdat %>% 
  mutate(triedbutnotdied = tried-deaths)
# In R missing values are typically NA values. A handy base function to know is is.na(). Try typing ?is.na in the console.  Let's use this function in a new formula to test if a variable is populated
witchdat <- witchdat %>% 
  mutate(missingyear =ifelse(is.na(year),"Yup missing","Not missing"))
# View the structure of witchdat again to see these new variables
str(witchdat, give.attr=FALSE)
```

##  Filter a "column"

In Excel we can filter by columns. Which functions can we use in R?

```{r filter}
# Filter which witches were tried in decade 1520 by using dplyr filter. Remember to use == for equals in R, whereas = is an assignment in R
witchdat %>% 
  filter(decade==1520) %>% 
  head()
```

##  Calculate on "columns"

In Excel we can add a calculation such as sum or a mean to a column. Which functions can we use in R?

```{r calculation}
# Sum the total deaths. Remember there are NA values so we need to tell this sum function to ignore these with argument na.rm=TRUE.Using sum on the column deaths is intuitively the same way of calculating this sum in Excel
sum(witchdat$deaths,na.rm=TRUE)
# Now try the dplyr summarise function
witchdat %>% 
  summarise(sum_deaths = sum(deaths, na.rm = TRUE))
```

##  Create a "Pivot Table"

In Excel you can create pivot tables. Which functions can we use in R?

A neat R package that creates interactive pivot tables is [rpivotTable](https://cran.r-project.org/web/packages/rpivotTable/).   Have a look at this package's [vignette](https://cran.r-project.org/web/packages/rpivotTable/vignettes/rpivotTableIntroduction.html).

```{r pivot}
# Create an interactive pivot table using the rpivotTable function and package
rpivotTable(witchdat,rows="country",vals="tried",aggregator="Sum")
```


## Save the "spreadsheet"

In Excel you can save your workbooks and spreadsheets. Which functions can we use in R?

Here are some options in R, depending on what output is needed:  

- The base R write.csv function to write to a csv file  
- The readr package in the tidyverse, includes the function write_csv, which is faster than write.csv  
- The xlsx package write.xlsx to write to an Excel workbook  


```{r save}
# Save dataframe with write_csv from readr package
write_csv(witchdat,"witchdat.csv")
```
  
  
  
***  

## Acknowledgments

  
Thanks to Marcia Ferreira @DrMarciaFe and Chris Beltz @BeltzEcology for reviewing and providing feedback on this tutorial.

