---
title: "Tutorial on Pivot Tables and other Excel things you can also do in R with Summer of Tech event data"
author: "Kim Fitter"
output: 
  html_document: 
    toc: true
    toc_float: true
---

## Introduction

This tutorial is aimed at those with a background in Excel, who would also like to use R for data analysis. This tutorial compares the things you would normally do in Excel, but done in R.

This introductory level tutorial assumes you have already installed R and R studio and had a brief introduction to the R basics and R Markdown. 

The dataset used is a real dataset of events run by a local non-profit Summer of Tech.

Run each line of code (Ctrl + Enter) or code chunk (Ctrl + Shift +Enter) yourself or Knit -> Knit to HTML to read through the HTML version.


```{r setup, include=FALSE}
# This is the setup chunk, we are using echo=TRUE so that the code is included in the R Markdown document
knitr::opts_chunk$set(cache = TRUE, echo = TRUE)
```

## Particularly useful packages for this exercise

```{r load packages}
# Load packages
library(tidyverse)
library(rpivotTable)
```

**Tip** Install the package from CRAN the first time you use the package using  `install.packages("name of package")` else you will get an `Error` that there is no package.

**Tip** The latest package version is sometimes available from GitHub so alternatively install from GitHub using the `devtools` package. A package's GitHub README file typically has installation instructions.


## About Summer of Tech

[Summer of Tech](https://summeroftech.co.nz/) (SoT) is a non-profit organisation in New Zealand that connects employers with students and graduates for paid work experience and graduate jobs. Students create a profile to register interest in a number of available summer internships. The students then attend skills and recruitment events, and SoT facilitates the whole process to job offers and securing these internships.

## Import Event data

The dataset used in this tutorial is Summer of Tech 2017 event details. 

First import the data using the read_csv function from the [readr](https://cran.r-project.org/web/packages/readr/index.html R package)   R package which is loaded with the tidyverse, to import the raw data to a data frame called `eventdat`.

A data frame is a 2-dimensional object where contents can be different data types.

```{r download data}
# Import the "2017 event detail report.csv" which is the "Raw" link from the GitHub link https://github.com/R-LadiesAKL/sotdata/blob/master/2017%20event%20detail%20report.csv
url <- "https://raw.githubusercontent.com/R-LadiesAKL/sotdata/master/2017%20event%20detail%20report.csv"
# The read_csv function also produces default messages describing the parsing of columns which is converting the columns into different types.
eventdat <- read_csv(url)
```

Let's take a quick look at the `eventdat` data frame to get an idea of it's size and contents. 

In R there are some packages that are included by default; take a look at the packages available in the Global Environment drop down. One such package is called [base](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html) which includes the `dim` function to retrieve the dimensions on the object: the row and column numbers.

```{r dim}
# Retrieve the number of rows and columns
dim(eventdat)
```

Another included package is [utils](https://cran.r-project.org/web/packages/R.utils/index.html) R package which includes the `str` and `structure` functions to view the basic structure. 

```{r structure}
# Use str function to 'Compactly Display the Structure of an Arbitrary R Object'. All R objects store additional attributes to store metadata about an object. In this str function we set the give.attr to FALSE to exclude these attributes.
str(eventdat, give.attr=FALSE)
# Next we will use structure function to view witchdat 
structure(eventdat)
# There are other ways of looking at your data. Try the glimpse function from the tibble package. Take a look at the Help description with ?glimpse to find out more
glimpse(eventdat)
```

This is a fairly small data frame with character, data/time, numeric columns. `Name` also appears to be the specific event name and a subset of the event `Type`.

##  Format "cells" in R

*In Excel we can format "cells". Which functions can we use in R?*

From now on we will refer to the columns as variables, with column headers as the variable names.

First we will look at the format of the variable names. 

```{r make names}
# Take a look at the column or variable names
names(eventdat)
#  We will use the make.names base R function to make syntactically valid names of the column variables for further analysis and output in R Markdown. For these names, it adds a "." to the spaces in the names
names(eventdat) <- make.names(names(eventdat))
```
  
A tool for data frame data manipulation is the  [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) R package. R packages typically have vignettes which are tutorials and worked examples using the package. There is useful documentation in the [Introduction to dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html).

In order to access the contents of the columns we can use a `$` for example `eventdat$Name`. Alternatively we can use another method that joins objects and functions using a pipe or `%>%`. This can be useful if we would like to use multiple `dplyr` verbs and functions on an object. 

```{r format}
# Use dplyr to rename the Name variable to Event.name
eventdat <- eventdat %>% 
  rename(Event.name=Name)
# Perform data type conversion of the deaths variable - use dplyr mutate to change this variable
eventdat <- eventdat %>% 
  mutate(Capacity=as.numeric(Capacity))
# or we can use dplyr mutate_if to change all integer variables (is.integer) to numeric variables (as.numeric). With mutate_if we don't need the () with is.integer and as.numeric functions as mutate_if will operate on all the columns for which the first test function is TRUE.
eventdat <- eventdat %>% 
  mutate_if(is.integer,as.numeric)
```

Try typing 'as.' in the Console to see what other function options are available to change formats.

```{r formatting changes structure}
# View the structure of eventdat again to see these formatting changes
glimpse(eventdat)
```

For other formatting we can also format the end products (such as plots), using the functions available in those R packages.

##  "Sort" data in R

*In Excel we can sort by columns. Which functions can we use in R?*

R is different to Excel that we can also view data manipulations without assigning changes to an object. In these sorting examples we will just view the changes in-line.

```{r arrange}
# Sort by character using arrange from dplyr. We can also pipe other packages functions such as head from the utils package.
eventdat %>% 
  arrange(Type) %>% 
  head()
# This also works on numeric variables, sort by year using arrange from dplyr
eventdat %>% 
  arrange(Capacity) %>% 
  head()
# Or we can arrange in descending order
eventdat %>% 
  arrange(desc(Capacity)) %>% 
  head()
```

##  Create a "formula" in R

*In Excel we can create formulas in cells. Which functions can we use in R?*

Remember R as a calculator? We can create a simple formula in R such as `3+2` but we can also create formula on parts or whole columns or cells as most R functions are vectorised. This is the same concept as filled cells in Excel.

```{r formula}
# Create a new variable in our dataframe called Extra.seats as the difference between two variables (formula after the =). Use the mutate function from dplyr
eventdat <- eventdat %>% 
  mutate(Extra.seats = Capacity - Number.registered)
# Take a look at the new variable and arrange in descending order
eventdat %>% 
  arrange(desc(Extra.seats)) %>% 
  head()
```

##  Filter a "column"

*In Excel we can filter by columns. Which functions can we use in R?*

```{r filter}
# Filter 100 Capacity events  by using dplyr filter. Remember to use == for equals in R, whereas = is an assignment operator in R functions.
eventdat %>% 
  filter(Capacity==100) %>% 
  head()
```

##  Calculate on "columns"

*In Excel we can add a calculation such as sum or a mean to a column. Which functions can we use in R?*

In R Markdown documents we can "highlight" and comment on insights, calculations or data manipulations.

```{r calculation}
# Sum the total events. Use sum on the 'whole column' ie variable eventdat$Number.checked.in is intuitively the same way of calculating this sum in Excel 
sum(eventdat$Number.checked.in)
# Now use dplyr with the select function to select the deaths variable (also the same as deleting the other columns in Excel) then apply the sum on this object
eventdat %>% 
  select(Number.checked.in) %>% 
  sum()
# Now try the dplyr summarise function and sum function then name this sum object sum_event
eventdat %>% 
  summarise(sum_event = sum(Number.checked.in))
```

##  Create a "Pivot Table"

*In Excel you can create pivot tables. Which functions can we use in R?*

In R we can use the table function to create a simple pivot table.

```{r table}
# Use th table function to find a simple count by country
eventdat %>% 
  select(Location) %>% 
      table()
```

A neat R package that creates interactive pivot tables is [rpivotTable](https://cran.r-project.org/web/packages/rpivotTable/).   Have a look at this package's [introduction vignette](https://cran.r-project.org/web/packages/rpivotTable/vignettes/rpivotTableIntroduction.html).

We will use the `rpivotTable` function and add arguments to specify how we would like the pivot table to look. We will use the event `Type` as a row and `Location` as Column, with a sum of the `Extra.seats`. The type of graphic will be a heatmap.

```{r pivot}
# Create an interactive pivot table heatmap using the rpivotTable function and package
rpivotTable(eventdat,rows=c("Type"),cols=c("Location"), aggregator="Sum", vals="Extra.seats", rendererName = "Heatmap")
```
  
In the heatmap object we can filter the `Location` and even add the `Event.name` as a row to see which specific events we over or under`Capacity`. We can also change the summation variable to one of the other numeric variables.

**Challenge 1** Can you recreate the pivot table as a new R data frame using `dplyr` functions? (Hint ?group_by)

**Challenge 2** On the pivot table widget move the country above the table so the `country` is a pivot table column. Can you extend your challenge 1 answer with `dplyr`  to recreate.

## Save the "spreadsheet"

*In Excel you can save your workbooks and spreadsheets. Which functions can we use in R?*

Here are some options in R, depending on what output is needed:  

- The `base` R `write.csv` function to write to a csv file  
- The `readr` package in the tidyverse, includes the function `write_csv`, which is faster than write.csv  
- The `xlsx` package `write.xlsx` to write to an Excel workbook  


```{r save}
# Save dataframe with write_csv from readr package
write_csv(eventdat,"eventdat.csv")
```


***  

## Acknowledgments
  

  
Thanks to Marcia Ferreira @DrMarciaFe, Chris Beltz @BeltzEcology, Duncan Garmonsway @nacnudus and Ivan Leung @urganmax for reviewing and providing suggesions on this tutorial through Twitter.
